---
title: "Lyme stats project code update"
author: "Kane Moser"
format: pdf
editor: visual
---

## [Plan]{.underline}

Here are the variables/parameters we want to have:

$X_{t,s}$: Lyme disease annual case incidence per state

$\Delta X_{t,s} = X_{t,s} - X_{t-1,s}$: delta X or change in incidence per state from year to year

$r_{t,s} = \frac{\Delta X_{t,s}}{X_{t-1,s}}$: rate of change in incidence per state from year to year

$\overline{r}_s = \frac{1}{14} \sum{r_{t,s}}$: mean rate of incidence change per state (this will be calculated from 2008-2021 data only)

$\tilde{X}_{t,s} = X_{t,s} \overline{r}_s$: predicted annual incidence based on previous step

$k_s = \frac{X_{t,s} - \tilde{X}_{t,s}}{\tilde{X}_{t,s}}$: difference between actual and predicted incidence (discrepancy presumably caused by change in case def.

Goal is to compare the distribution of $k_s$ for high and low incidence areas to then be able to test some simple hypotheses.

Possible hypotheses to test:

1.  We expect the mean of the distribution of $k_s$ for low incidence states to be centered on 0.
2.  We expect the mean of the distribution of $k_s$ for high incidence states to *not* be centered on 0.
3.  We expect the mean of the distribution of $k_s$ for high and low incidence states to be different from each other.

### 1. Load packages and clean data:

```{r load data}
library(tidyverse)

data <- read_csv("data/Lyme_Disease_Incidence_Rates_by_State_or_Locality.csv")
# Remove special characters from "State" column
data$State <- str_remove_all(data$State, "[^[:alnum:] ]")
# Put data in long format
data <- data %>% pivot_longer(!State, names_to = "Year", values_to = "Incidence")

jurisdiction_data <- read_csv("data/Lyme_jurisdiction_data.csv") %>%
  rename("State" = "states")

data <- data %>%
  left_join(jurisdiction_data, by="State") %>%
  select(State,Year,Incidence,jurisdiction) %>%
  filter(Incidence > 0)
```

### 2. Calculate variable values

```{r}
# calculate delta_X
data <- data %>%
  group_by(State) %>%
  mutate(X = Incidence) %>%
  mutate(delta_X = X - lag(X)) %>%
  ungroup()

# calculate r_t
data <- data %>%
  group_by(State) %>%
  mutate(r = delta_X / lag(X)) %>%
  ungroup()

# calculate mean rate of incidence change per state, mean_r
mean_rates <- data %>%
  filter(Year >= 2008, Year <= 2021) %>% # filter years for 2008-2021
  group_by(State) %>%
  summarise(mean_r = mean(r, na.rm = TRUE))

# join mean rates back to main data
data <- data %>%
  left_join(mean_rates, by = "State")
```

```{r}
# Calculate X_pred and k
# predicted value for 2022, X_pred
state_data <- data %>%
  group_by(State) %>%
  mutate(X_pred = ifelse(Year == 2022, lag(X) * mean_r, NA)) %>%
  ungroup() %>%
  na.omit()

# difference between predicted value and actual value, k
state_data <- state_data %>%
  mutate(k = (X - X_pred) / X_pred) %>%
  filter(!is.infinite(k), !is.na(k))
```

### Compare the distribution of k for low and high incidence states

```{r}
p1 <- ggplot(state_data, aes(x=k, fill = jurisdiction)) +
  geom_histogram(color = "grey40", position="dodge") +
  #facet_wrap(~jurisdiction) +
  theme_minimal() +
  scale_fill_viridis_d()
p1
```

### Hypothesis testing

```{r}
low <- state_data %>% filter(jurisdiction == "low")
high <- state_data %>% filter(jurisdiction == "high")
summary(low)
summary(high)

t.test(low$k, mu=0)
```

```{r}
t.test(high$k, mu = 0)
```

```{r}
t.test(k ~ jurisdiction, state_data)
```
